Laplacian envelope

### 2021-11-22 Notes/laplace.tex
unrearranged a bit
degeneracy section

We can almost see that degeneracy implies lumpability as follows:
suppose there is a nonzero transition between states \(i\) and \(j\) of type \(\nu\),
Then \(\kktm^\nu_{ij} = 0\) and \cref{eq:areaderivs} becomes
%
\begin{equation}\label{eq:dgn_theta}
\begin{gathered}
  \pdiff{\CL}{\MMdm^\nu_{ij}} =
    r f^\nu \eqm_i \prn{ \brk{\thbm_i - \thbm_j}
     + \prn{c_i + (-1)^\nu + \lambda} \brk{\etwm_i - \etwm_j} } 
    = 0, \\
    \implies \quad
    r f^\nu \eqm_i = 0
    \qquad \text{or} \qquad
    \thbm_i - \thbm_j = -  \prn{c_i + (-1)^\nu + \lambda} \brk{\etwm_i - \etwm_j}. 
\end{gathered}
\end{equation}
%
Therefore, provided that state \(i\) has nonzero equilibrium probability, degeneracy in \(\etw\) implies degeneracy in \(\thb\).

Let \(C_a\) denote the \(a\)'th clump of states that are degenerate in \(\etw\) and \(\thb\), in order of decreasing \(\etw\):

@@

subsection{Diagonal constraints}\label{sec:diagKKT}

\Cref{eq:KKTz} will provide another expression for \(\kktm^\nu_{mm}\).
That expression will agree with \cref{eq:diagKKTz} provided that
%
\begin{equation}\label{eq:diagAgree}
    \hat{A} - c_m (\delta \wm_m + s \delta \etwm_m)
      - \lambda \delta w_m 
      = (\dgnm_{m,m} - \dgnm_{m*}) 
        + \Delta f (\etwm_m - \etwm_{m*}).
\end{equation}
%

@@

>>>>

We can almost see that degeneracy implies lumpability as follows:
suppose there is a nonzero transition between states \(i\) and \(j\) of type \(\nu\),
Then \(\kktm^\nu_{ij} = 0\) and \cref{eq:areaderivs} becomes
%
\begin{equation}\label{eq:dgn_theta}
\begin{gathered}
  \pdiff{\CL}{\MMdm^\nu_{ij}} =
    r f^\nu \eqm_i \prn{ \brk{\thbm_i - \thbm_j}
     + \prn{c_i + (-1)^\nu + \lambda} \brk{\etwm_i - \etwm_j} } 
    = 0, \\
    \implies \quad
    r f^\nu \eqm_i = 0
    \qquad \text{or} \qquad
    \thbm_i - \thbm_j = -  \prn{c_i + (-1)^\nu + \lambda} \brk{\etwm_i - \etwm_j}. 
\end{gathered}
\end{equation}
%
Therefore, provided that state \(i\) has nonzero equilibrium probability, degeneracy in \(\etw\) implies degeneracy in \(\thb\) for an otimal model.

Let \(C_a\) denote the \(a\)'th clump of states that are degenerate in \(\etw\) and \(\thb\), in order of decreasing \(\etw\).
The arguments from \cref{sec:areapert} tell us that the transitions in \(\W\pot\) from states in \(C_a\) can only go to states in \(C_{a+1}\).%
\footnote{We haven't said anything about intra-clump transitions yet.
Perturbing such transitions does not break the degeneracy, so these perturbations can be extended finitely.
Those perturbations also leave the area invariant, so they are completely null degrees of freedom.
We can choose to set all intra-clump transitions to zero.}
Similarly, \(\W\dep\) can only go from states in \(C_a\) to states in \(C_{a-1}\).

Suppose we have two states, \(i,j\), in \(C_a\), and two states, \(k,l\), in \(C_{a+1}\), with \(\Wm\pot_{ik}, \Wm\pot_{jl} > 0\).
Then \cref{eq:dgn_theta} tells us that
%
\begin{equation}\label{eq:dgn_c}
\begin{aligned}
  \thbm_i - \thbm_k &= - \prn{c_i + 1 + \lambda} \brk{\etwm_i - \etwm_k} 
  \quad \implies & 
  \thbm_a - \thbm_{a+1} &= - \prn{c_i + 1 + \lambda} \brk{\etwm_a - \etwm_{a+1}}, \\
  \thbm_j - \thbm_l &= - \prn{c_j + 1 + \lambda} \brk{\etwm_j - \etwm_l}
  \quad \implies & 
  \thbm_a - \thbm_{a+1} &= - \prn{c_j + 1 + \lambda} \brk{\etwm_a - \etwm_{a+1}}, \\
  && \implies \quad 
  0 &= \prn{c_i - c_j} \brk{\etwm_a - \etwm_{a+1}}. \\
\end{aligned}
\end{equation}
%
Therefore \(\mathbf{c}\) has the same degeneracy as \(\etw\) and \(\thb\).
If we make the choice \(\wm_i = \sgn c_i\) (as discussed after \cref{eq:lareacoeffs}) we see that \(\w\) also has the same degeneracy.
A result similar to \cref{eq:dgn_c} holds with \( \text{pot} \leftrightarrow \text{dep} \).

Note that, for any state \(i \in C_a\)
%
\begin{equation*}
\begin{aligned}
  [\W\pot \etw]_i &= \sum_{j \in C_{a+1}} \Wm\pot_{ij} (\etwm_j - \etwm_i)
    = (\etwm_{a+1} - \etwm_a) \sum_{\mathclap{j \in C_{a+1}}} \Wm\pot_{ij}
    = (\etwm_{a+1} - \etwm_a) \lambda\pot_i, \\
  [\W\dep \etw]_i &= (\etwm_{a-1} - \etwm_a) \lambda\dep_i,
  \quad\text{where} \quad
  \lambda^\nu_i =  \sum_{j \neq i} \Wm^\nu_{ij}.
\end{aligned}
\end{equation*}
%
which also applies wth \(\etw\) replaced with \(\thb\) or \(\mathbf{c}\).
All models, not just the optimal ones, satisfy
%
\begin{equation*}
\begin{aligned}
  \frg \thb(0) &= (\I - \onev \eq) \mathbf{a} = - \enc \etw(0) - \CA,
  \\
  \frg \etw(0) &= (\I - \onev \eq) \w = \w - \Delta p\, \onev.
\end{aligned}
\end{equation*}
%
Using the results above
%
\begin{equation}\label{eq:dgn_lump}
\begin{gathered}
\begin{aligned}
  (\thbm_{a+1} - \thbm_a) \lambda\pot_i + (\thbm_{a-1} - \thbm_a) \lambda\dep_i
    &= - (\etwm_{a+1} - \etwm_a) \lambda\pot_i + (\etwm_{a-1} - \etwm_a) \lambda\dep_i - \CA,
  \\
  (\etwm_{a+1} - \etwm_a) \lambda\pot_i + (\etwm_{a-1} - \etwm_a) \lambda\dep_i &= \wm_a - \Delta p,
\end{aligned}
\\
  \implies
\begin{aligned}
  \lambda\pot_i &= \frac{(\CA - \wm_a + \Delta p) (\etwm_a - \etwm_{a-1}) 
                        + (\wm_a - \Delta p) (\thbm_a - \thbm_{a-1})}
                      {(\etwm_{a+1} - \etwm_a) (\thbm_a - \thbm_{a-1})
                        - (\etwm_a - \etwm_{a-1}) (\thbm_{a+1} - \thbm_a)
                        - 2 (\etwm_{a+1} - \etwm_a) (\etwm_a - \etwm_{a-1})}, \\
  \lambda\dep_i &= \frac{(\CA + \wm_a - \Delta p) (\etwm_{a+1} - \etwm_a) 
                        + (\wm_a - \Delta p) (\thbm_{a+1} - \thbm_a)}
                      {(\etwm_{a+1} - \etwm_a) (\thbm_a - \thbm_{a-1})
                        - (\etwm_a - \etwm_{a-1}) (\thbm_{a+1} - \thbm_a)
                        - 2 (\etwm_{a+1} - \etwm_a) (\etwm_a - \etwm_{a-1})}.
\end{aligned}
\end{gathered}
\end{equation}
%
Because the right-hand sides above depend on \(a\) and not \(i\), \cref{eq:dgn_lump} implies that \(\boldsymbol{\lambda}^\nu\) also has the same degeneracy as \(\etw\), which is the condition for lumpability under the partitions \(\brc{C_a}\) \cite{kemeny1960finite,Ball1993Lumpability,burke1958markovian}.

However, if we substitute the last of \cref{eq:dgn_theta} into \cref{eq:dgn_lump} we find that the denominators vanish and the numerators reduce to \( [\CA - (c_a + \lambda) (\wm_a - \Delta p)] (\etwm_a - \etwm_{a-1}) \), etc.
When the numerator vanishes, the first two equations of \cref{eq:dgn_lump} are degenerate rather than contradictory, leaving one linear combination of the \(\lambda^\nu_i\) undetermined.
This means we must have \( \CA = (c_a + \lambda) (\wm_a - \Delta p) \), which would seem to imply that there are only two allowed values of \(c_a\).

This bizarre situation can be avoided if we note the other possibility in \cref{eq:dgn_theta}: we could have \(\eqm_i = 0\).
As we will see below in \cref{eq:serialextarea}, the optimal models have \(\eqm_i = 0\) for all but two states.
Using this alternative loses the degeneracy of \(\thb\) and \(\w\), leaving us just short of proving lumpability.

@@

@@

\subsection{Diagonal constraints}\label{sec:diagKKT}

\Cref{eq:KKTz} provided another expression for \(\kktm^\nu_{mm}\).
That expression will agree with \cref{eq:diagKKTz} provided that
%
\begin{equation}\label{eq:diagAgree}
    \hat{A} - c_m (\delta \wm_m + s \delta \etwm_m)
      - \lambda \delta w_m 
      = (\dgnm_{m,m} - \dgnm_{m*}) 
        + \Delta f (\etwm_m - \etwm_{m*}).
\end{equation}
%
